Shift-sensitive Power Transformations for Transforming Data to Normality

### 2023-01-10 manuscript/manuscript.Rmd
WIP on Figure 1.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

@@

Numerical features in datasets may strongly deviate from normal distributions, e.g. by being skewed. This may complicate further analysis. Power transformations [@Tukey1957-rt] can help improve normality of such features. The two most commonly used transformations are that of @Box1964-mz and @Yeo2000-vw. The Box-Cox transformation of a feature $x$ under the transformation parameter $\lambda$ is defined as:

@@

```{=latex}
\begin{equation}
\phi_{\text{YJ}}^\lambda (x) = 
\begin{cases}
\left( \left( 1 + x \right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x \geq 0\\
\log(1 + x) & \text{if } \lambda = 0 \text{ and } x \geq 0\\
-\left( \left( 1 - x\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x < 0\\
-\log(1 - x) & \text{if } \lambda = 2 \text{ and } x < 0
\end{cases}
\end{equation}
```
The $\lambda$-parameter is typically optimised using maximum likelihood optimisation under the assumption that the transformed feature is normally distributed. As noted by Raymaekers and Rousseeuw, this approach is very sensitive to the presence of outliers, and robust versions of Box-Cox and Yeo-Johnson transformations were devised [@Raymaekers2021-kq].

Power transformation does not always yield transformed variables that

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  fig.align= "center")

library(power.transform)

# Base font size.
base_size <- 10

get_annotation_settings <- function(ggtheme=NULL){
  # Import formatting settings from the provided ggtheme.

  # Find the text size for the table. This is based on text sizes in the
  # ggtheme.
  fontsize <- ggtheme$text$size
  fontsize_rel <- 1.0

  # Attempt to base the text size on the general axis.text attribute.
  if(!is.null(ggtheme$axis.text$size)){
    if(inherits(ggtheme$axis.text$size, "rel")){
      # Find the relative text size of axis text.
      fontsize_rel <- as.numeric(ggtheme$axis.text$size)

    } else {
      # Set absolute text size.
      fontsize <- ggtheme$axis.text$size
      fontsize_rel <- 1.0
    }
  }

  # Attempt to refine the text size using the axis.text.y attribute in
  # particular.
  if(!is.null(ggtheme$axis.text.y$size)){
    if(inherits(ggtheme$axis.text.y$size, "rel")){
      # Set relative text size of axis text
      fontsize_rel <- as.numeric(ggtheme$axis.text.y$size)

    } else {
      # Set absolute text size.
      fontsize <- as.numeric(ggtheme$axis.text.y$size)
      fontsize_rel <- 1.0
    }
  }

  # Update the text size using the magical ggplot2 point size (ggplot2:::.pt).
  geom_text_size <- fontsize * fontsize_rel / 2.845276

  # Obtain lineheight
  lineheight <- ggtheme$text$lineheight
  if(!is.null(ggtheme$axis.text$lineheight)) lineheight <- ggtheme$axis.text$lineheight
  if(!is.null(ggtheme$axis.text.y$lineheight)) lineheight <- ggtheme$axis.text.y$lineheight

  # Obtain family
  fontfamily <- ggtheme$text$family
  if(!is.null(ggtheme$axis.text$family)) fontfamily <- ggtheme$axis.text$family
  if(!is.null(ggtheme$axis.text.y$family)) fontfamily <- ggtheme$axis.text.y$family
  if(!is.null(ggtheme$axis.text.x$family)) fontfamily <- ggtheme$axis.text.x$family

  # Obtain face
  fontface <- ggtheme$text$face
  if(!is.null(ggtheme$axis.text$face)) fontface <- ggtheme$axis.text$face
  if(!is.null(ggtheme$axis.text.y$face)) fontface <- ggtheme$axis.text.y$face
  if(!is.null(ggtheme$axis.text.x$face)) fontface <- ggtheme$axis.text.x$face

  # Obtain colour
  colour <- ggtheme$text$colour
  if(!is.null(ggtheme$axis.text$colour)) colour <- ggtheme$axis.text$colour
  if(!is.null(ggtheme$axis.text.y$colour)) colour <- ggtheme$axis.text.y$colour
  if(!is.null(ggtheme$axis.text.x$colour)) colour <- ggtheme$axis.text.x$colour

  return(list(
    "geom_text_size"=geom_text_size,
    "fontsize"=fontsize,
    "fontsize_rel"=fontsize_rel,
    "colour"=colour,
    "family"=fontfamily,
    "face"=fontface,
    "lineheight"=lineheight))
}

```

@@

Numerical variables in datasets may strongly deviate from normal distributions, e.g. by being skewed. This may complicate further analysis. Power transformations [@Tukey1957-rt] can help improve normality of such features. The two most commonly used transformations are that of @Box1964-mz and @Yeo2000-vw. The Box-Cox transformation of a feature $x$ under the transformation parameter $\lambda$ is defined as:

@@

```{=latex}
\begin{equation}
\phi_{\text{YJ}}^\lambda (x) = 
\begin{cases}
\left( \left( 1 + x \right)^\lambda - 1\right) / \lambda & \text{if } \lambda \neq 0 \text{ and } x \geq 0\\
\log(1 + x) & \text{if } \lambda = 0 \text{ and } x \geq 0\\
-\left( \left( 1 - x\right)^{2 - \lambda} - 1 \right) / \left(2 - \lambda \right) & \text{if } \lambda \neq 2 \text{ and } x < 0\\
-\log(1 - x) & \text{if } \lambda = 2 \text{ and } x < 0
\end{cases}
\end{equation}
```
The $\lambda$-parameter is typically optimised using maximum likelihood estimation under the assumption that the transformed feature is normally distributed. As noted by Raymaekers and Rousseeuw, this approach is very sensitive to the presence of outliers, and robust versions of Box-Cox and Yeo-Johnson transformations were devised [@Raymaekers2021-kq].

Power transformation does not guarantee that transformed variables are normally distributed. In fact, depending on location and scale of the variable, power transformation may reduce normality, as shown in Figure \@ref(fig:reduced-normality).

``` {r reduced-normality, echo="FALSE", include="FALSE", fig.cap="Power transformations may reduce normality depending on location. Variable $x$ consists of $10000$ values are randomly drawn from a normal distribution $\\mathcal{N}(\\mu, 1)$. For both Box-Cox and Yeo-Johnson transformations, we would expect $\\lambda = 1$, since $x$ is already normally distributed. Panel A and B show the $\\lambda$ value found through maximum likelihood estimation for Box-Cox and Yeo-Johnson transformations, respectively."}

compute_lambda <- function(mu, x, method){
  # Create transformer object.
  transformer <- suppressWarnings(
    power.transform::find_transformation_parameters(
      x = x + mu,
      method = method,
      robust = FALSE,
      shift = FALSE,
      lambda = NULL))

  # Return lambda value.
  return(transformer@lambda)
}

x <- stats::rnorm(10000L)
shift_range <- 10^seq(from=0, to=6, by=0.1)

box_cox_values <- sapply(
  shift_range,
  compute_lambda,
  x = x,
  method = "box_cox")

yeo_johnson_values <- sapply(
  shift_range,
  compute_lambda,
  x = x,
  method = "yeo_johnson")

data <- data.table::data.table(
  "mu" = log10(c(shift_range, shift_range)),
  "lambda" = c(box_cox_values, yeo_johnson_values),
  "method" = factor(x=rep(c("Box-Cox", "Yeo-Johnson"), each = length(shift_range), levels=c("Box-Cox", "Yeo-Johnson"))))

annotation_settings <- get_annotation_settings(ggplot2::theme_light(base_size = base_size))

p <- ggplot2::ggplot(
  data = data,
  mapping = ggplot2::aes(
    x = mu,
    y = lambda,
    colour = method))
p <- p + ggplot2::theme_light(base_size=base_size)
p <- p + ggplot2::geom_point()
p <- p + ggplot2::geom_hline(yintercept = 1.0, linetype="longdash", colour="grey40")
p <- p + ggplot2::annotate(
  geom = "text",
  x = 6,
  y = 1,
  label = "expected",
  colour=annotation_settings$colour,
  family=annotation_settings$family,
  fontface=annotation_settings$face,
  size=annotation_settings$geom_text_size,
  vjust=1.5,
  hjust=1.0)
p <- p + ggplot2::scale_x_continuous(labels = scales::math_format())

p
```
